Inference methods for extended latent Gaussian models

### 2022-11-02 src/docs_15-11-22-seminar/15-11-22-seminar.Rmd
Iterate seminar a little (e.g. include AGHQ plots)

* Consider areas $i = 1, \ldots, n$
* Simple random sample household-survey taken in each area, with sample sizes $m^\text{HS}_i$
* The number of people testing positive is $y^\text{HS}_i$
* Then we can use a binomial logistic regression of the form:
\begin{align*}
y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
\text{logit}(\rho^\text{HS}_i) &\sim g(\vartheta^\text{HS}), \quad i = 1, \ldots, n,
\end{align*}
* If $g$ is Gaussian then this is a latent Gaussian model in the sense of @rue2009approximate
* We usually set up $g$ as a spatial smoother, because the sample sizes in each area are too small to get reliable direct estimates
* One problem with the above model is that household surveys are expensive to run, so they only happen rarely

@@

* So what is `TMB`?
* R package which implements the Laplace approximation for latent variable models using automatic differentiation (via `CppAD`)
  * For more about AD see e.g. @griewank2008evaluating
  * Useful for getting the mode, Hessian
* Write an objective function $f(\x, \btheta)$ in C++ ("user template")
    * We select $f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$

@@

* Suggested reading: the paper [@rue2009approximate] or a book e.g. @blangiardo2015spatial
* Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which recall are three-stage models with middle layer
\begin{equation*}
  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
\end{equation*}
* `R-INLA` implementation takes advantage of sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)

@@

2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
  * For low $m$ `R-INLA` uses a grid-strategy (illustrated in the next slide)
  * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design is used
  * Other approaches, like adaptive Gaussian Hermite quadrature (AGHQ) have recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and may be preferable

@@

* `aghq` R package and vignette [@stringer2021implementing]
* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior)
* Use $k$ quadrature points
  * If $k$ is odd then they include the mode
  * If $k = 1$ then it's a Laplace approximation
  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often

>>>>

* Consider small-areas $i = 1, \ldots, n$
* Simple random sample household-survey in each area, with sample sizes $m^\text{HS}_i$
* The number of people testing positive is $y^\text{HS}_i$
* Then we can use a binomial logistic regression of the form:
\begin{align*}
y^\text{HS}_i &\sim \text{Bin}(m^\text{HS}_i, \rho^\text{HS}_i), \\
\text{logit}(\rho^\text{HS}_i) &\sim g(\vartheta^\text{HS}), \quad i = 1, \ldots, n,
\end{align*}
* If $g$ is Gaussian then this is a latent Gaussian model in the sense of @rue2009approximate
* We usually set up $g$ as a spatial smoother, because the sample sizes in each area are too small to get reliable direct estimates
* One problem with the above model is that household surveys are expensive to run, so they only happen rarely

@@

* R package which implements the Laplace approximation for latent variable models using automatic differentiation (via `CppAD`)
  * For more about AD see e.g. @griewank2008evaluating. It's useful for getting the mode, Hessian
* Write an objective function $f(\x, \btheta)$ in C++ ("user template")
    * We select $f(\x, \btheta) = - \log p(\y \, | \, \x, \btheta) p(\x \, | \, \btheta) p(\btheta)$

@@

* Suggested reading: the paper @rue2009approximate or a book e.g. @blangiardo2015spatial
* Approximate Bayesian inference for \textcolor{hilit}{latent Gaussian models} (LGMs), which recall are three-stage models with middle layer
\begin{equation*}
  \text{(Latent field)} \qquad  p(\x \, | \, \btheta) = \mathcal{N}(\x  \, | \, \bm{\mu}(\btheta), \bm{Q}(\btheta)^{-1}).
\end{equation*}
* `R-INLA` implementation takes advantage of sparsity properties of $\bm{Q}(\btheta)$, i.e. if $\x$ is a Gaussian Markov random field (GMRF)

@@

2) In both \eqref{eq:inla1} and \eqref{eq:inla2} we want to integrate w.r.t. \eqref{eq:hypermarginal}, so choose integration points and weights $\{ \btheta^{(k)}, \Delta^{(k)} \}$
  * For low $m$ `R-INLA` uses a grid-strategy
  * For larger $m$ this becomes too expensive and `R-INLA` uses a CCD design
  * Other approaches, like adaptive Gaussian Hermite quadrature (AGHQ) have recently been shown to have theoretical guarantees [@bilodeau2021stochastic] and may be preferable

# Grid strategy

* To write...

@@

* Gauss-Hermite quadrature is a way of picking nodes and weights, and is based on the theory of polynomial interpolation
* The adaptive part means that it uses the location (mode) and curvature (Hessian) of the target (posterior) to automatically choose the node locations
  * No manual tuning!
* Works particularly well when the integrand is pretty Gaussian
* Supposing you use $k$ quadrature points, then
  * If $k = 1$ then it's a Laplace approximation
  * In the vignette $k = 3$ (for each dimension, so $3^m$ total) is chosen quite often
* Implemented in the `aghq` R package. See vignette @stringer2021implementing

```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("One dimensional example of AGHQ. \\label{fig:aghq-1d}")}
knitr::include_graphics("depends/aghq-1d.pdf")
```

```{r message=FALSE, echo=FALSE, out.height="200px", fig.cap=paste("Two dimensional example of AGHQ. \\label{fig:aghq-2d}")}
knitr::include_graphics("depends/aghq-2d.pdf")
```
